Static and Dynamic Bethe-Salpeter Equations in the $T$-Matrix Approximation

### 2021-11-03 Manuscript/Tmatrix.tex
abstract

\begin{abstract}
We report, for the first time, the static and dynamic Bethe-Salpeter equations (BSE) within the so-called $T$-matrix approximation (BSE@$GT$).
The performance of the (static) BSE@$GT$ scheme and its perturbative dynamical correction (dBSE@$GT$) are assessed by computed the neutral excited states of molecular systems.
%\bigskip
%\begin{center}
%	\boxed{\includegraphics[width=0.5\linewidth]{TOC}}
%\end{center}
%\bigskip
\end{abstract}

@@

In order to target neutral excited states, we first consider the static version of the BSE employing the $GT$ quasiparticle energies [see Eq.~\eqref{eq:G0T0}] as well as $T$-matrix kernel [\ie, $\Sigma^{\GT}$ in Eq.~\eqref{eq:Xi}].
In this case, the BSE@$GT$ linear eigenvalue problem reads
\begin{equation}
\label{eq:BSE}
	\begin{pmatrix}
		\bA{}{\BSE}		&	\bB{}{\BSE}	\\
		-\bB{}{\BSE}	&	-\bA{}{\BSE}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{m}{\BSE}	\\
		\bY{m}{\BSE}	\\
	\end{pmatrix}
	=
	\Om{m}{\BSE}
	\begin{pmatrix}
		\bX{m}{\BSE}	\\
		\bY{m}{\BSE}	\\
	\end{pmatrix}
\end{equation}
with 
\begin{subequations}
\begin{align}
	A_{ia,jb}^{\BSE} & = \delta_{ij} \delta_{ab} (\e{a}{\GT} - \e{i}{\GT}) + \mel{ib}{}{aj} + T_{ij,ba}(\omega=0)
	\\ 
	B_{ia,jb}^{\BSE} & = \mel{ij}{}{ab} + T_{ib,ja}(\omega=0)
\end{align}
\end{subequations}
The eigenvalues of Eq.~\eqref{eq:BSE} provide $OV$ singlet (\ie, spin-conserved) and $OV$ triplet  (\ie, spin-flip) single excitations.
Due to the frequency-independent nature of static BSE, it is well known that one cannot access double (and higher) excitations.
\cite{Loos_2019,Romaniello_2009b,Sangalli_2011,Loos_2020h,Authier_2020,Monino_2021}

In order to go beyond the static approximation, it is possible to consider the dynamical version of the BSE (dBSE) within the Tamm-Dancoff approximation. \cite{Strinati_1988}
In this case, one must solve the (non-linear) dynamical eigenvalue problem
\begin{equation}
\label{eq:dBSE}
	\begin{pmatrix}
		\bA{}{\dBSE}(\Om{S}{})	&	\bB{}{\BSE}	\\
		-\bB{}{\BSE}	&	-\bA{}{\dBSE}(-\Om{S}{})	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
	=
	\Om{S}{}
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
\end{equation}
with 
\begin{equation}
	A_{ia,jb}^{\dBSE}(\omega) = \delta_{ij} \delta_{ab} (\e{a}{\GT} - \e{i}{\GT}) + \mel{ib}{}{aj} + \widetilde{T}_{ij,ba}(\omega)
\end{equation}
where, following Strinati's seminal work, \cite{Strinati_1988} one can derive the following expression for the elements of the dynamical $T$ matrix
\begin{equation}
\label{eq:dT}
\begin{split}
	\widetilde{T}_{pq,rs}(\omega) 
		& = \sum_{n} \frac{\braket*{pq}{\chi_n^{N+2}}\braket*{rs}{\chi_n^{N+2}}}{\omega - \Om{n}{N+2} + (\e{i}{\GT} + \e{j}{\GT}) + i\eta}
		\\
		& + \sum_{n} \frac{\braket*{pq}{\chi_n^{N-2}}\braket*{rs}{\chi_n^{N-2}}}{\omega + \Om{n}{N-2} - (\e{a}{\GT} + \e{b}{\GT}) + i\eta}
\end{split}
\end{equation}
Equation \eqref{eq:dT} highlights the interesting dynamical structure of the $T$ matrix, where, similarly to the dBSE@$GW$ scheme, \cite{Strinati_1988,Romaniello_2009b,Loos_2020h} the 2h2p configurations are downfolded on the 1h1p configurations.

@@

The present formalism has been implemented in QuAcK \cite{QuAcK} and we consider here only systems with closed-shell singlet ground states.
The $GW$ and $GT$ calculations are performed by considering a (restricted) HF starting point and standard gaussian basis sets (defined with cartesian functions) are employed.

>>>>

\begin{abstract}
While the well-established $GW$ approximation corresponds to a resummation of the direct ring diagrams and is particularly well suited for weakly-correlated systems, the $T$-matrix approximation does sum ladder diagrams up to infinity and is supposedly more appropriate in the presence of strong correlation.
Here, we derive and implement, for the first time, the static and dynamic Bethe-Salpeter equations when one consider $T$-matrix quasiparticle energies as well as a $T$-matrix-based kernel.
The performance of the static scheme and its perturbative dynamical correction are assessed by computed the neutral excited states of molecular systems.
Comparison with more conventional schemes as well as other wave function methods are also performed.
%\bigskip
%\begin{center}
%	\boxed{\includegraphics[width=0.5\linewidth]{TOC}}
%\end{center}
%\bigskip
\end{abstract}

@@

In order to target neutral excited states, we first consider the static version of the BSE employing the $GT$ quasiparticle energies [see Eq.~\eqref{eq:G0T0}] as well as $T$-matrix kernel [\ie, $\Sigma^{\GT}$ in Eq.~\eqref{eq:Xi}].
In this case, the BSE@$GT$ linear eigenvalue problem reads
\begin{equation}
\label{eq:BSE}
	\begin{pmatrix}
		\bA{}{\BSE}		&	\bB{}{\BSE}	\\
		-\bB{}{\BSE}	&	-\bA{}{\BSE}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{m}{\BSE}	\\
		\bY{m}{\BSE}	\\
	\end{pmatrix}
	=
	\Om{m}{\BSE}
	\begin{pmatrix}
		\bX{m}{\BSE}	\\
		\bY{m}{\BSE}	\\
	\end{pmatrix}
\end{equation}
with 
\begin{subequations}
\begin{align}
	A_{ia,jb}^{\BSE} & = \delta_{ij} \delta_{ab} (\e{a}{\GT} - \e{i}{\GT}) + \mel{ib}{}{aj} + T_{ij,ba}(\omega=0)
	\\ 
	B_{ia,jb}^{\BSE} & = \mel{ij}{}{ab} + T_{ib,ja}(\omega=0)
\end{align}
\end{subequations}
The eigenvalues of Eq.~\eqref{eq:BSE} provide $OV$ singlet (\ie, spin-conserved) and $OV$ triplet  (\ie, spin-flip) single excitations.
Note that the spin structure of the BSE@$GT$ equations are completely analogous to the BSE@$GW$ version. \cite{Monino_2021}
Therefore, one can compute separately singlet and triplet excitation energies.

Due to the frequency-independent nature of static BSE, it is well known that one cannot access double (and higher) excitations.
\cite{Loos_2019,Romaniello_2009b,Sangalli_2011,Loos_2020h,Authier_2020,Monino_2021}
In order to go beyond the static approximation, it is possible to consider the dynamical version of the BSE (dBSE) within the Tamm-Dancoff approximation. \cite{Strinati_1988}
In this case, one must solve the (non-linear) dynamical eigenvalue problem
\begin{equation}
\label{eq:dBSE}
	\begin{pmatrix}
		\bA{}{\dBSE}(\Om{S}{})	&	\bB{}{\BSE}	\\
		-\bB{}{\BSE}	&	-\bA{}{\dBSE}(-\Om{S}{})	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
	=
	\Om{S}{}
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
\end{equation}
with 
\begin{equation}
	A_{ia,jb}^{\dBSE}(\omega) = \delta_{ij} \delta_{ab} (\e{a}{\GT} - \e{i}{\GT}) + \mel{ib}{}{aj} + \widetilde{T}_{ij,ba}(\omega)
\end{equation}
where, following Strinati's seminal work, \cite{Strinati_1988} one can derive the following expression for the elements of the dynamical $T$ matrix
\begin{equation}
\label{eq:dT}
\begin{split}
	\widetilde{T}_{pq,rs}(\omega) 
		& = \sum_{n} \frac{\braket*{pq}{\chi_n^{N+2}}\braket*{rs}{\chi_n^{N+2}}}{\omega - \Om{n}{N+2} + (\e{i}{\GT} + \e{j}{\GT}) + i\eta}
		\\
		& + \sum_{n} \frac{\braket*{pq}{\chi_n^{N-2}}\braket*{rs}{\chi_n^{N-2}}}{\omega + \Om{n}{N-2} - (\e{a}{\GT} + \e{b}{\GT}) + i\eta}
\end{split}
\end{equation}
Equation \eqref{eq:dT} highlights the interesting dynamical structure of the $T$ matrix, where, similarly to the dBSE@$GW$ scheme, \cite{Strinati_1988,Romaniello_2009b,Loos_2020h} the 2h2p configurations are downfolded on the 1h1p configurations.

@@

The present formalism has been implemented in QuAcK \cite{QuAcK} and we consider here only systems with closed-shell singlet ground states.
The $GW$ and $GT$ calculations are performed by considering a (restricted) HF starting point and standard gaussian basis sets (defined with cartesian functions) are employed.
\titou{Discuss the computational cost of the BSE@$GT$ equations.}
